---
title: "Palmer Penguins Initial Analysis"
author: "Adaobi Nwankwo"
format: html
editor: visual
execute:
  keep-md: true
---

## Palmer Penguins Analysis

This is an analysis of the Palmer's Penguin dataset.

## Loading Packages and Datasets

Here we will load the tidyverse package and penguins data.

```{r}
#Load the tidyverse
library(tidyverse)
library(kableExtra)

#Read the penguins_samp1 data file from github
penguins <- read_csv("https://raw.githubusercontent.com/mcduryea/Intro-to-Bioinformatics/main/data/penguins_samp1.csv")

#See the first six rows of the data we've read in to our notebook
penguins %>%
  head() %>%
  kable() %>%
  kable_styling(c("hover", "striped"))
```

## About Our Data

The data we are working with is a data set on Penguins, which includes 8 features measured on 44 Penguins. The features that are included are physiological features (like bill length, bill depth, flipper length, body mass, etc.) as well as other features, like the year the Penguin was observed, the Island the Penguin was observed on, the sex of the Penguin, and the species of the Penguin.

## Interesting Questions to Ask

-   What is the average flipper length? What about for each species?

-   Are there more male or female Penguins? What about per island or species?

-   What is the average body mass? What about by Island? By species? By sex?

-   What is the ratio of bill length to bill depth for a Penguin? What is the overall average of this metric? Does it change by species, sex, or island?

-   Does average body mass change by year?

## Data Manipulation Tools and Strategies

We can look at individual columns in a data set or subsets of columns in a data set. For example, if we are only interested in flipper length and species, we can `select()` those column.

```{r}
penguins %>%
  select(species, body_mass_g)

```

If we want to `filter()` and only show certain rows, we can do that too. Here, we filtered by sex, numerical variables, and both.

```{r}
#we can filter by sex (categorical variables)
penguins %>%
  filter(species == "chinstrap")

#we can also filter by numerical variables
penguins %>%
  filter(body_mass_g >= 6000)

#we can also do both
penguins %>%
  filter((body_mass_g >= 6000) | (island == "Torgersen"))
  

```

## Answering Our Questions

Most of our questions involve summarizing data, and perhaps summarizing over groups. We can summarize data using the `summarize()` function, and group data using `group_by()`.

Let's find the average flipper length.

Using this function, I found that the Gentoo had the highest average flipper length of 218 and the Adelie had the lowest average flipper length of 189.

```{r}
#overall average flipper length
penguins %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))

#single species average 
penguins %>%
  filter(species == "Gentoo") %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))

#grouped average
penguins %>%
  group_by(species) %>%
  summarize(avg_flipper_length = mean (flipper_length_mm))

  
```

How many of each species do we have? By using this function, I found that Gentoo had the highest count of species. This might mean that the data for the Gentoo species is a bit more accurate than the data for the Chinstrap species because the Chinstrap species has a count of 2. Since there's only 2 of them it would be harder ti fund an accurate average.

```{r}
penguins %>%
  count(species)

```

Are there more males of females penguins? In the whole data set, it was found that there are more male species. Using different functions allows for a more specific finding when comparing sex in the data set. I was able to group sex by species and island and this was an interesting find.

```{r}
penguins %>%
  count(sex)
penguins %>%
  group_by(species) %>%
  count(sex)
penguins %>%
  group_by(island) %>%
  count(sex)


```

We can use a function called mutate to add new columns to our data set.

```{r}
penguins %>%
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm)

```

```{r}
penguins_with_ratio <- penguins %>% 
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm)

#average ratio
penguins %>%
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm) %>%
  summarize(mean_bill_ltd_ratio = mean(bill_ltd_ratio), 
            median_bill_ltd_ratio = median(bill_ltd_ratio))
            

```

Average body mass by year.

We can use the `summarize` function to find the average body mass by year. The average body mass was highest in year 2007 with a body mass of 5,078.

```{r}
penguins %>%
  group_by(year) %>%
  summarize(mean_body_mass = mean(body_mass_g))

```

## Data Visualization

-   What is the distribution of penguin flipper length?

-   What is the distribution of penguin species?

-   Does the distribution of flipper length depend on the species of the penguin?

The next 3 graphs:

The next few graphs contain ggplots. In the first plot, we compared species to their count. We did this earlier, but here we can visualize it which makes it a bit easier to compare. In the second plot we measured the distribution of bill lengths. Here it was found that the majority of the bill lengths had a length around 45mm. For the third plot we measured the distribution of flipper length. We also added a vline which showed us the mean and the median of the data. Again, we had this information before but here we look at it in a visualize way.

```{r}
penguins %>%
  ggplot() +
  geom_bar(mapping = aes(x = species)) +
  labs(title ="Counts of Penguin Species", x = "Species", y = "Count")

```

```{r}
penguins %>%
  ggplot() +
  geom_histogram(mapping = aes(x = bill_length_mm),
                 color = "white",
                 fill = "blue") +
  labs(title ="Distribution of Bill Lengths",
       x = "Bill Length (mm)", y = "Count")

```

```{r}
penguins %>%
  ggplot() +
  geom_histogram(mapping = aes(x = flipper_length_mm),
                 bins = 15,
                 fill = "forestgreen",
                 color = "black") +
  labs(title = "Distribution of Flipper Length (mm)",
       #subtitle: "mean in black, median in purple"
       y = "", x = "Flipper Length (mm)") +
  geom_vline(aes(xintercept = mean(flipper_length_mm)), lwd = 2, lty = "dashed") +
  geom_vline(aes(xintercept = median(flipper_length_mm)), color = "purple", lwd = 2,   lty = "dotted")
  
       
```

Let's make a scatter plot to see if bill length is correlated with bill depth. This scatter plot allowed us to compare bill length and bill depth by species. Given that the Chinstrap only had 2 counts, it's scatterplot is very bland while the Gentoos have many points. When looking at this scatterplot it's evident that bill length and bill depth are correlated. As the bill depth increases, the bill length increases.

```{r}
penguins %>%
  ggplot() +
  geom_point(mapping=aes(x = bill_depth_mm, y = bill_length_mm, color = species)) +
  geom_smooth(aes(x = bill_depth_mm, y = bill_length_mm, color = species), method = "lm") +
  labs(title = "Bill Depth and Length",
       x = "Bill Depth (mm)",
       y = "Bill Length (mm)")

```

We will now look at the distribution of species. Here we learned how to use color which was a cool tool.

```{r}
penguins %>%
  ggplot() +
  geom_bar(mapping = aes(x = species), color = "purple", fill = "pink") +
  labs(title = "Counts of Penguin Species",
       x = "Species",
       y = "count")

```

We also learned how to add and title x and y intercepts and titles. With this data, it is evident that the Island Biscoe had the most species while Dream Island had the fewest species.

```{r}
penguins %>%
  ggplot() +
  geom_bar(mapping = aes(x = island, fill = species)) +
  labs(title = "Species by Island",
       x = "Island",
       y = "Species")

```

Here we use a ggplot to look at the bill depth by species. In specific we can see the median bill depth for each species. We can see that the median bill depth varies for each species. In the plot under that we can see the mean for each species, this varies as well for each species. In the last plot, we compare flipper and bill lengths by species and sex. This also varies and there doesn't seem to be a correlation.

```{r}
penguins %>%
  ggplot() +
  geom_boxplot(mapping = aes(x = bill_depth_mm, y = species)) +
  labs(title = "Bill Depth by Species",
       x = "Bill Depth (mm)",
       y = "Median")
```

```{r}
penguins %>%
  ggplot() +
  geom_histogram(mapping = aes(x = bill_depth_mm)) +
  facet_wrap(~species, ncol = 1) +
  labs(title = "Bill Depth by Species",
       x = "Bill Depth (mm)",
       y = "mean")
```

```{r}
penguins %>%
  filter(!is.na(sex)) %>%
  ggplot() +
  geom_point(mapping = aes(x = bill_length_mm,
                           y = flipper_length_mm,
                           color = species,
                           shape = sex)) +
  geom_smooth(mapping = aes(x = bill_length_mm,
                            y = flipper_length_mm,
                            color = species),
              method = "lm") +
  facet_grid(island ~ sex, scales = "free") +
  labs(title = "Flipper and Bill Lengths by Species and Sex",
       x = "Bill Length (mm)",
       y = "Flipper Length (mm)") +
  theme_light()
  
```

## A Final Question

Determine whether the average bill length for a penguin exceeds 45 mm. It was found that the average bill length for a penguin is about 46.3 mm so it does exceed 45 mm.

```{r}
penguins %>%
  summarize(avg_bill_length = mean(bill_length_mm))

t.test(penguins$bill_length_mm, alternative = "greater", mu = 45, conf.level = 0.95)

```
