---
title: "Palmer Penguins Initial Analysis"
author: "Adaobi Nwankwo"
format: html
editor: visual
execute:
  keep-md: true
---

## Palmer Penguins Analysis

This is an analysis of the Palmer's Penguin dataset.

## Loading Packages and Datasets

Here we will load the tidyverse package and penguins data.

```{r}
#Load the tidyverse
library(tidyverse)
library(kableExtra)

#Read the penguins_samp1 data file from github
penguins <- read_csv("https://raw.githubusercontent.com/mcduryea/Intro-to-Bioinformatics/main/data/penguins_samp1.csv")

#See the first six rows of the data we've read in to our notebook
penguins %>%
  head() %>%
  kable() %>%
  kable_styling(c("hover", "striped"))
```

## About Our Data

The data we are working with is a data set on Penguins, which includes 8 features measured on 44 Penguins. The features that are included are physiological features (like bill length, bill depth, flipper length, body mass, etc.) as well as other features, like the year the Penguin was observed, the Island the Penguin was observed on, the sex of the Penguin, and the species of the Penguin.

## Interesting Questions to Ask

-   What is the average flipper length? What about for each species?

-   Are there more male or female Penguins? What about per island or species?

-   What is the average body mass? What about by Island? By species? By sex?

-   What is the ratio of bill length to bill depth for a Penguin? What is the overall average of this metric? Does it change by species, sex, or island?

-   Does average body mass change by year?

## Data Manipulation Tools and Strategies

We can look at individual columns in a data set or subsets of columns in a data set. For example, if we are only interested in flipper length and species, we can `select()` those column.

```{r}
penguins %>%
  select(species, body_mass_g)

```

If we want to `filter()` and only show certain rows, we can do that too.

```{r}
#we can filter by sex (categorical variables)
penguins %>%
  filter(species == "chinstrap")

#we can also filter by numerical variables
penguins %>%
  filter(body_mass_g >= 6000)

#we can also do both
penguins %>%
  filter((body_mass_g >= 6000) | (island == "Torgersen"))
  

```

## Answering Our Questions

Most of our questions involve summarizing data, and perhaps summarizing over groups. We can summarize data using the `summarize()` function, and group data using `group_by()`.

Let's find the average flipper length.

```{r}
#overall average flipper length
penguins %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))

#single species average 
penguins %>%
  filter(species == "Gentoo") %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))

#grouped average
penguins %>%
  group_by(species) %>%
  summarize(avg_flipper_length = mean (flipper_length_mm))

  
```

How many of each species do we have?

```{r}
penguins %>%
  count(species)

```

Are there more males of females penguins?

```{r}
penguins %>%
  count(sex)
penguins %>%
  group_by(species) %>%
  count(sex)
penguins %>%
  group_by(island) %>%
  count(sex)


```

We can use a function called mutate to add new columns to our data set.

```{r}
penguins %>%
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm)

```

```{r}
penguins_with_ratio <- penguins %>% 
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm)

#average ratio
penguins %>%
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm) %>%
  summarize(mean_bill_ltd_ratio = mean(bill_ltd_ratio), 
            median_bill_ltd_ratio = median(bill_ltd_ratio))
            

```

average body mass by year

```{r}
penguins %>%
  group_by(year) %>%
  summarize(mean_body_mass = mean(body_mass_g))

```

## Data Visualization

-   What is the distribution of penguin flipper length?

-   What is the distribution of penguin species?

-   Does the distribution of flipper length depend on the species of the penguin?

```{r}
penguins %>%
  ggplot() +
  geom_bar(mapping = aes(x = species)) +
  labs(title ="Counts of Penguin Species", x = "Species", y = "Count")

```

```{r}
penguins %>%
  ggplot() +
  geom_histogram(mapping = aes(x = bill_length_mm),
                 color = "white",
                 fill = "blue") +
  labs(title ="Distribution of Bill Lengths",
       x = "Bill Length (mm)", y = "Count")

```

```{r}
penguins %>%
  ggplot() +
  geom_histogram(mapping = aes(x = flipper_length_mm),
                 bins = 15,
                 fill = "forestgreen",
                 color = "black") +
  labs(title = "Distribution of Flipper Length (mm)",
       #subtitle: "mean in black, median in purple"
       y = "", x = "Flipper Length (mm)") +
  geom_vline(aes(xintercept = mean(flipper_length_mm)), lwd = 2, lty = "dashed") +
  geom_vline(aes(xintercept = median(flipper_length_mm)), color = "purple", lwd = 2,   lty = "dotted")
  
       
```

Let's make a scatter plot to see if bill length is correlated with bill depth.

```{r}
penguins %>%
  ggplot() +
  geom_point(mapping=aes(x = bill_depth_mm, y = bill_length_mm, color = species)) +
  geom_smooth(aes(x = bill_depth_mm, y = bill_length_mm, color = species), method = "lm") +
  labs(title = "Bill Depth and Length",
       x = "Bill Depth (mm)",
       y = "Bill Length (mm)")

```

We will now look at the distribution of species.

```{r}
penguins %>%
  ggplot() +
  geom_bar(mapping = aes(x = species), color = "purple", fill = "pink") +
  labs(title = "Counts of Penguin Species",
       x = "Species",
       y = "count")

```

```{r}
penguins %>%
  ggplot() +
  geom_bar(mapping = aes(x = island, fill = species)) +
  labs(title = "Species by Island",
       x = "Island",
       y = "Species")

```

```{r}
penguins %>%
  ggplot() +
  geom_boxplot(mapping = aes(x = bill_depth_mm, y = species)) +
  labs(title = "Bill Depth by Species",
       x = "Bill Depth (mm)",
       y = "Median")
```

```{r}
penguins %>%
  ggplot() +
  geom_histogram(mapping = aes(x = bill_depth_mm)) +
  facet_wrap(~species, ncol = 1) +
  labs(title = "Bill Depth by Species",
       x = "Bill Depth (mm)",
       y = "mean")
```

```{r}
penguins %>%
  filter(!is.na(sex)) %>%
  ggplot() +
  geom_point(mapping = aes(x = bill_length_mm,
                           y = flipper_length_mm,
                           color = species,
                           shape = sex)) +
  geom_smooth(mapping = aes(x = bill_length_mm,
                            y = flipper_length_mm,
                            color = species),
              method = "lm") +
  facet_grid(island ~ sex, scales = "free") +
  labs(title = "Flipper and Bill Lengths by Species and Sex",
       x = "Bill Length (mm)",
       y = "Flipper Length (mm)") +
  theme_light()
  
```

## A Final Question

Determine whether the average bill length for a penguin exceeds 45 mm.

```{r}
penguins %>%
  summarize(avg_bill_length = mean(bill_length_mm))

t.test(penguins$bill_length_mm, alternative = "greater", mu = 45, conf.level = 0.95)

```
